---
title: "choropleth"
author: "Markus Lundsfryd Jensen"
date: "2026-05-25"
output: html_document
---

```{r}
library(pacman)
pacman::p_load("tidyverse","sf","rnaturalearth","rnaturalearthdata","patchwork", "ggspatial")
```


# Set parameters

## Same bin boundaries as KDE script
```{r}
CRISIS_START <- as.Date("2005-09-30")
CRISIS_END   <- as.Date("2006-03-30")
CORPUS_START <- as.Date("2005-03-30")
CORPUS_END   <- as.Date("2006-09-30")
```


## Minimum mentions per country to be included in the difference map
```{r}
MIN_COUNTRY_MENTIONS <- 1
```


# Map extent (same crop as KDE for visual consistency)
```{r}
MAP_XLIM <- c(-20, 80)
MAP_YLIM <- c(10, 72)
```


#Load data
```{r}
mentions <- read_csv("../out/place_mentions.csv") %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= CORPUS_START, date <= CORPUS_END) %>%
  filter(!is.na(country)) %>%
  mutate(period = case_when(
    date < CRISIS_START                        ~ "pre_crisis",
    date >= CRISIS_START & date <= CRISIS_END  ~ "crisis_peak",
    date > CRISIS_END                          ~ "post_crisis"
  )) %>%
  mutate(period = factor(period,
    levels = c("pre_crisis", "crisis_peak", "post_crisis")))
```


# Normalise by total mentions per period so counts are comparable
# Raw counts would conflate geographic expansion with higher article volume
```{r}
period_totals <- mentions %>%
  count(period, name = "total_mentions")

country_counts <- mentions %>%
  count(country, period) %>%
  left_join(period_totals, by = "period") %>%
  mutate(mention_share = n / total_mentions)
```


# Wide format: one row per country, columns for each period
```{r}
country_wide <- country_counts %>%
  select(country, period, mention_share) %>%
  pivot_wider(
    names_from  = period,
    values_from = mention_share,
    values_fill = 0
  )
```



# Difference scores (crisis peak relative to pre-crisis baseline)
```{r}
country_wide <- country_wide %>%
  mutate(
    diff_crisis   = crisis_peak - pre_crisis,
    diff_recovery = post_crisis - crisis_peak
  )

# Apply minimum mention filter across all periods combined (legacy)
country_mention_totals <- mentions %>%
  count(country, name = "total") %>%
  filter(total >= MIN_COUNTRY_MENTIONS)
  

#Define dictionrary for translation
danish_to_natural_earth <- c(
  # A
  "Afghanistan"                                                                    = "Afghanistan",
  "Albanien"                                                                       = "Albania",
  "Algeriet"                                                                       = "Algeria",
  "Amerikas Forenede Stater"                                                       = "United States",
  "Andorra"                                                                        = "Andorra",
  "Angola"                                                                         = "Angola",
  "Anguilla"                                                                       = NA,
  "Antigua og Barbuda"                                                             = "Antigua and Barbuda",
  "Argentina"                                                                      = "Argentina",
  "Armenien"                                                                       = "Armenia",
  "Aserbajdsjan"                                                                   = "Azerbaijan",
  "Australien"                                                                     = "Australia",
  # B
  "Bahamas"                                                                        = "Bahamas",
  "Bahrain"                                                                        = "Bahrain",
  "Bangladesh"                                                                     = "Bangladesh",
  "Barbados"                                                                       = "Barbados",
  "Belgien"                                                                        = "Belgium",
  "Belize"                                                                         = "Belize",
  "Benin"                                                                          = "Benin",
  "Bermuda"                                                                        = NA,
  "Bhutan"                                                                         = "Bhutan",
  "Bolivia"                                                                        = "Bolivia",
  "Bosnien og Hercegovina"                                                         = "Bosnia and Herzegovina.",
  "Botswana"                                                                       = "Botswana",
  "Brasilien"                                                                      = "Brazil",
  "Britiske Indiske Ocean"                                                         = NA,
  "Brunei"                                                                         = "Brunei Darussalam",
  "Bulgarien"                                                                      = "Bulgaria",
  "Burkina Faso"                                                                   = "Burkina Faso",
  "Burundi"                                                                        = "Burundi",
  # C
  "Cambodia"                                                                       = "Cambodia",
  "Cambodja"                                                                       = "Cambodia",
  "Cameroun"                                                                       = "Cameroon",
  "Canada"                                                                         = "Canada",
  "Centralafrikanske Republik"                                                     = "Central African Republic",
  "Chile"                                                                          = "Chile",
  "Colombia"                                                                       = "Colombia",
  "Costa Rica"                                                                     = "Costa Rica",
  "Cuba"                                                                           = "Cuba",
  "Cypern"                                                                         = "Cyprus",
  # D
  "Danmark"                                                                        = "Denmark",
  "De palæstinensiske områder"                                                     = "Palestine",
  "Demokratiske Republik Congo"                                                    = "Republic of the Congo",
  "Den Islamiske Republik Iran"                                                    = "Iran",
  "Djibouti"                                                                       = "Djibouti",
  "Dominica"                                                                       = "Dominica",
  "Dominikanske Republik"                                                          = "Dominican Republic",
  # E
  "East Germany"                                                                   = NA,
  "Ecuador"                                                                        = "Ecuador",
  "Egypten"                                                                        = "Egypt",
  "El Salvador"                                                                    = "El Salvador",
  "Elfenbenskysten"                                                                = "Côte d'Ivoire",
  "Eritrea"                                                                        = "Eritrea",
  "Estland"                                                                        = "Estonia",
  "Etiopien"                                                                       = "Ethiopia",
  # F
  "Fiji"                                                                           = "Fiji",
  "Filippinerne"                                                                   = "Philippines",
  "Finland"                                                                        = "Finland",
  "Forenede Arabiske Emirater"                                                     = "United Arab Emirates",
  "France"                                                                         = "France",
  "Frankrig"                                                                       = "France",
  "Færøerne"                                                                       = "Faeroe Is.",
  # G
  "Gabon"                                                                          = "Gabon",
  "Gambia"                                                                         = "The Gambia",
  "Georgien"                                                                       = "Georgia",
  "Ghana"                                                                          = "Ghana",
  "Gibraltar"                                                                      = NA,
  "Greece"                                                                         = "Greece",
  "Grækenland"                                                                     = "Greece",
  "Grønland"                                                                       = "Greenland",
  "Guatemala"                                                                      = "Guatemala",
  "Guernsey"                                                                       = NA,
  "Guinea"                                                                         = "Guinea",
  "Guinea-Bissau"                                                                  = "Guinea-Bissau",
  "Guyana"                                                                         = "Guyana",
  # H
  "Haiti"                                                                          = "Haiti",
  "Holland"                                                                        = "Netherlands",
  "Honduras"                                                                       = "Honduras",
  "Hviderusland"                                                                   = "Belarus",
  # I
  "Indien"                                                                         = "India",
  "Indonesia"                                                                      = "Indonesia",
  "Indonesien"                                                                     = "Indonesia",
  "Irak"                                                                           = "Iraq",
  "Iran"                                                                           = "Iran",
  "Ireland"                                                                        = "Ireland",
  "Irland"                                                                         = "Ireland",
  "Island"                                                                         = "Iceland",
  "Israel"                                                                         = "Israel",
  "Italien"                                                                        = "Italy",
  # J
  "Jamaica"                                                                        = "Jamaica",
  "Japan"                                                                          = "Japan",
  "Jordan"                                                                         = "Jordan",
  # K
  "Kap Verde"                                                                      = "Cape Verde",
  "Kasakhstan"                                                                     = "Kazakhstan",
  "Kenya"                                                                          = "Kenya",
  "Kina"                                                                           = "China",
  "Kirgisistan"                                                                    = "Kyrgyzstan",
  "Kosovo"                                                                         = "Kosovo",
  "Kroatien"                                                                       = "Croatia",
  "Kuwait"                                                                         = "Kuwait",
  # L
  "Laos"                                                                           = "Lao PDR",
  "Lebanon"                                                                        = "Lebanon",
  "Letland"                                                                        = "Latvia",
  "Libanon"                                                                        = "Lebanon",
  "Liberia"                                                                        = "Liberia",
  "Libyen"                                                                         = "Libya",
  "Liechtenstein"                                                                  = "Liechtenstein",
  "Litauen"                                                                        = "Lithuania",
  "Luxembourg"                                                                     = "Luxembourg",
  # M
  "Madagaskar"                                                                     = "Madagascar",
  "Malawi"                                                                         = "Malawi",
  "Malaysia"                                                                       = "Malaysia",
  "Mali"                                                                           = "Mali",
  "Malta"                                                                          = "Malta",
  "Marokko"                                                                        = "Morocco",
  "Marshalløerne"                                                                  = "Marshall Is.",
  "Mauretanien"                                                                    = "Mauritania",
  "Mauritius"                                                                      = "Mauritius",
  "Mexico"                                                                         = "Mexico",
  "Moldova"                                                                        = "Moldova",
  "Monaco"                                                                         = "Monaco",
  "Mongoliet"                                                                      = "Mongolia",
  "Montenegro"                                                                     = "Montenegro",
  "Mozambique"                                                                     = "Mozambique",
  "Moçambique"                                                                     = "Mozambique",
  "Myanmar"                                                                        = "Myanmar",
  # N
  "Namibia"                                                                        = "Namibia",
  "Nederlandene"                                                                   = "Netherlands",
  "Nepal"                                                                          = "Nepal",
  "New Zealand"                                                                    = "New Zealand",
  "Nicaragua"                                                                      = "Nicaragua",
  "Niger"                                                                          = "Niger",
  "Nigeria"                                                                        = "Nigeria",
  "Nordcypern / Cypern"                                                            = "N. Cyprus",
  "Nordkorea"                                                                      = "Dem. Rep. Korea",
  "Nordmakedonien"                                                                 = "North Macedonia",
  "Norge"                                                                          = "Norway",
  # O
  "Oman"                                                                           = "Oman",
  # P
  "Pakistan"                                                                       = "Pakistan",
  "Panama"                                                                         = "Panama",
  "Papua Ny Guinea"                                                                = "Papua New Guinea",
  "Paraguay"                                                                       = "Paraguay",
  "Pavestolen (Vatikanstaten)"                                                     = "Vatican",
  "Peru"                                                                           = "Peru",
  "Philippines"                                                                    = "Philippines",
  "Polen"                                                                          = "Poland",
  "Portugal"                                                                       = "Portugal",
  # Q
  "Qatar"                                                                          = "Qatar",
  # R
  "Republikken Congo"                                                              = "Republic of the Congo",
  "Congo"                                                              = "Republic of the Congo",
  "Republikken Korea (Sydkorea)"                                                   = "Republic of Korea",
  "República Árabe Saharaui Democrática الجمهورية العربية الصحراوية الديمقراطية"  = "W. Sahara",
  "Rumænien"                                                                       = "Romania",
  "Rusland"                                                                        = "Russian Federation",
  "Rwanda"                                                                         = "Rwanda",
  # S
  "Saint Vincent og Grenadinerne"                                                  = "St. Vin. and Gren.",
  "Salomonøerne"                                                                   = "Solomon Islands",
  "Samoa"                                                                          = "Samoa",
  "San Marino"                                                                     = "San Marino",
  "Saudi-Arabien"                                                                  = "Saudi Arabia",
  "Schweiz"                                                                        = "Switzerland",
  "Senegal"                                                                        = "Senegal",
  "Serbien"                                                                        = "Serbia",
  "Sierra Leone"                                                                   = "Sierra Leone",
  "Singapore"                                                                      = "Singapore",
  "Slovakiet"                                                                      = "Slovakia",
  "Slovenien"                                                                      = "Slovenia",
  "Somalia"                                                                        = "Somalia",
  "Soomaaliland أرض الصومال"                                                       = "Somaliland",
  "South Africa"                                                                   = "South Africa",
  "Spanien"                                                                        = "Spain",
  "Sri Lanka"                                                                      = "Sri Lanka",
  "Storbritannien"                                                                 = "United Kingdom",
  "Sudan"                                                                          = "Sudan",
  "Surinam"                                                                        = "Suriname",
  "Svalbard and Jan Mayen"                                                         = NA,
  "Sverige"                                                                        = "Sweden",
  "Sydafrika"                                                                      = "South Africa",
  "Sydkorea"                                                                       = "Republic of Korea",
  "Sydossetien"                                                                    = NA,
  "Sydsudan"                                                                       = "S. South Sudan",
  "Syria"                                                                          = "Syria",
  "Syrien"                                                                         = "Syria",
  # T
  "Taiwan"                                                                         = "Taiwan",
  "Tanzania"                                                                       = "Tanzania",
  "Tchad"                                                                          = "Chad",
  "Thailand"                                                                       = "Thailand",
  "Tjekkiet"                                                                       = "Czech Republic",
  "Togo"                                                                           = "Togo",
  "Tokelau"                                                                        = NA,
  "Tonga"                                                                          = "Tonga",
  "Trinidad og Tobago"                                                             = "Trinidad and Tobago",
  "Tunesien"                                                                       = "Tunisia",
  "Turkmenistan"                                                                   = "Turkmenistan",
  "Tyrkiet"                                                                        = "Turkey",
  "Tyskland"                                                                       = "Germany",
  # U
  "USA"                                                                            = "United States",
  "Uganda"                                                                         = "Uganda",
  "Ukraine"                                                                        = "Ukraine",
  "Ungarn"                                                                         = "Hungary",
  "United Kingdom"                                                                 = "United Kingdom",
  "Uruguay"                                                                        = "Uruguay",
  "Usbekistan"                                                                     = "Uzbekistan",
  # V
  "Vanuatu"                                                                        = "Vanuatu",
  "Vatikanstaten"                                                                  = "Vatican",
  "Venezuela"                                                                      = "Venezuela",
  "Vietnam"                                                                        = "Vietnam",
  # Y
  "Yemen"                                                                          = "Yemen",
  # Z
  "Zambia"                                                                         = "Zambia",
  "Zimbabwe"                                                                       = "Zimbabwe",
  # Æ Ø Å
  "Østrig"                                                                         = "Austria",
  "Østtimor"                                                                       = "Timor-Leste",
  # Arabic
  "الأراضي الفلسطينية"                                                             = "Palestine"
)

# Apply translation
country_wide <- country_wide %>%
  mutate(country = ifelse(
    country %in% names(danish_to_natural_earth),
    danish_to_natural_earth[country],
    country
  )) %>%
  filter(!is.na(country))  # drop territories with no polygon
```


# Load world polygons and join

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  select(name, name_long, iso_a3, geometry)

# Country name matching between OpenCage output and Natural Earth is imperfect
# Some manual fixes may be needed - check join_check below for unmatched countries
choropleth_data <- world %>%
  left_join(country_wide, by = c("name_long" = "country"))

# show which countries in the data didn't match polygons
join_check <- country_wide %>%
  filter(!country %in% world$name_long) %>%
  pull(country)

if (length(join_check) > 0) {
  cat("Countries in data not matched to polygons. may need manual name fixes:\n")
  print(join_check)
}
```
#OBS cannot fix these. SHould not be big problem, as they are outside the scope of europe and are eliminated for all periods


#functinof or plotting

```{r}
# Calculate symmetric limit spanning both difference rasters
shared_limit <- max(
  abs(country_wide$diff_crisis),
  abs(country_wide$diff_recovery),
  na.rm = TRUE
)

# Then pass it explicitly to both difference plots
plot_choropleth <- function(data, fill_var, title,
                            diverging = FALSE,
                            fixed_limit = NULL,
                            low = "#2c7bb6",
                            mid = "white",
                            high = "#d7191c") {

  fill_values <- data[[fill_var]]

  if (diverging) {
    limit <- if (!is.null(fixed_limit)) fixed_limit else max(abs(fill_values), na.rm = TRUE)
    scale <- scale_fill_gradient2(
      low      = low,
      mid      = mid,
      high     = high,
      midpoint = 0,
      limits   = c(-limit, limit),
      na.value = "grey85",
      name     = "Δ mention\nshare"
    )
  } else {
    scale <- scale_fill_gradient(
      low      = "white",
      high     = "#d73027",
      na.value = "grey85",
      name     = "Mention\nshare"
    )
  }

  ggplot(data) +
    geom_sf(aes(fill = .data[[fill_var]]), color = "grey40", linewidth = 0.2) +
    scale +
    coord_sf(xlim = MAP_XLIM, ylim = MAP_YLIM) +
    annotation_scale(
    location      = "br",
    width_hint    = 0.2,
    style         = "bar",
    unit_category = "metric"
  ) +
    labs(title = title) +
    theme_minimal() +
    theme(
      axis.title      = element_blank(),
      axis.text       = element_blank(),
      panel.grid      = element_line(color = "grey90"),
      legend.position = "right",
      plot.title      = element_text(size = 20)
    )
}
```



# Individual period choropleths

```{r}
p_pre <- plot_choropleth(
  choropleth_data, "pre_crisis",
  "Pre-crisis mention share"
)

p_peak <- plot_choropleth(
  choropleth_data, "crisis_peak",
  "Crisis peak mention share"
)

p_post <- plot_choropleth(
  choropleth_data, "post_crisis",
  "Post-crisis mention share"
)
```


# Difference choropleths

```{r}
# Pass shared_limit to both difference plots
p_diff_crisis <- plot_choropleth(
  choropleth_data, "diff_crisis",
  "Change in mention share: Crisis Peak vs Pre-Crisis",
  diverging    = TRUE,
  fixed_limit  = shared_limit
)

p_diff_recovery <- plot_choropleth(
  choropleth_data, "diff_recovery",
  "Change in mention share: Post-Crisis vs Crisis Peak",
  diverging    = TRUE,
  fixed_limit  = shared_limit
)
```


# Combined figures using patchwork

```{r}
# Three-period overview
p_overview <- p_pre / p_peak / p_post +
  plot_annotation(
    title    = "Geographic distribution of news coverage across periods",
    subtitle = "Mention share normalised within each period",
    theme    = theme(plot.title = element_text(size = 13, face = "bold"))
  )

# Difference maps side by side
p_diffs <- p_diff_crisis | p_diff_recovery +
  plot_annotation(
    title    = "Shifts in geographic attention before, during and after the cartoon crisis",
    subtitle = "Difference in normalised mention share between periods",
    theme    = theme(plot.title = element_text(size = 20, face = "bold"))
  )
```


# Save

```{r}
ggsave("../out/choropleth/choropleth_overview.png", p_overview,
       width = 10, height = 16, dpi = 300)

ggsave("../out/choropleth/choropleth_diffs.png", p_diffs,
       width = 16, height = 7,  dpi = 300)

# Also save individually
ggsave("../out/choropleth/choropleth_pre.png",           p_pre,           width = 10, height = 6, dpi = 300)
ggsave("../out/choropleth/choropleth_peak.png",          p_peak,          width = 10, height = 6, dpi = 300)
ggsave("../out/choropleth/choropleth_post.png",          p_post,          width = 10, height = 6, dpi = 300)
ggsave("../out/choropleth/choropleth_diff_crisis.png",   p_diff_crisis,   width = 10, height = 6, dpi = 300)
ggsave("../out/choropleth/choropleth_diff_recovery.png", p_diff_recovery, width = 10, height = 6, dpi = 300)

```

