---
title: "centroid combined"
author: "Markus Lundsfryd Jensen"
date: "2026-05-25"
output: html_document
---

```{r}
library(pacman)
pacman::p_load("tidyverse","sf","terra","rnaturalearth","rnaturalearthdata","patchwork", "ggspatial")
```


# ==============================================================================
# PARAMETERS
# ==============================================================================

```{r}
CRISIS_START <- as.Date("2005-09-30")
CRISIS_END   <- as.Date("2006-03-30")
CORPUS_START <- as.Date("2005-03-30")
CORPUS_END   <- as.Date("2006-09-30")

KDE_BANDWIDTH <- 2
RASTER_RES    <- 1
MIN_MENTIONS  <- 1
MAP_XLIM      <- c(-20, 80)
MAP_YLIM      <- c(10, 72)
```


```{r}
# Centroid computation method: "mean" or "weighted"
# "mean" treats every mention equally regardless of how often a place appears
# "weighted" weights by mention frequency, pulling centroid toward hotspots

CENTROID_METHOD <- "weighted"


```


#Load and prepare data

```{r}
mentions <- read_csv("../out/place_mentions.csv") %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= CORPUS_START, date <= CORPUS_END) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  filter(lat >= -90, lat <= 90, lon >= -180, lon <= 180) %>%
  mutate(period = case_when(
    date < CRISIS_START                        ~ "pre_crisis",
    date >= CRISIS_START & date <= CRISIS_END  ~ "crisis_peak",
    date > CRISIS_END                          ~ "post_crisis"
  )) %>%
  mutate(period = factor(period,
    levels = c("pre_crisis", "crisis_peak", "post_crisis")))

mention_counts <- mentions %>%
  count(place) %>%
  filter(n >= MIN_MENTIONS)

mentions <- mentions %>%
  filter(place %in% mention_counts$place)

cat("Mentions per period:\n")
print(table(mentions$period))
```

#COmpute centroids pr period
```{r}
compute_centroid <- function(df, method = "weighted") {

  if (method == "weighted") {
    # Weight each mention by how often that place appears in the period
    weights <- df %>% count(place, lon, lat, name = "w")
    centroid_lon <- weighted.mean(weights$lon, weights$w)
    centroid_lat <- weighted.mean(weights$lat, weights$w)
  } else {
    centroid_lon <- mean(df$lon)
    centroid_lat <- mean(df$lat)
  }

  tibble(lon = centroid_lon, lat = centroid_lat)
}

centroids <- mentions %>%
  group_by(period) %>%
  group_modify(~ compute_centroid(.x, method = CENTROID_METHOD)) %>%
  ungroup()
```


# Convert to sf for plotting
```{r}
centroids_sf <- centroids %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

cat("\nCentroids per period:\n")
print(centroids)
```


# Compute KDE for all periods, resample and save
```{r}
mentions_sf <- mentions %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

raster_template <- rast(
  xmin = -180, xmax = 180,
  ymin = -90,  ymax = 90,
  resolution = RASTER_RES,
  crs = "EPSG:4326"
)

compute_kde <- function(points_sf, template, bandwidth) {

  if (nrow(points_sf) < 3) {
    warning("Too few points for KDE - returning empty raster")
    r <- template
    values(r) <- 0
    return(r)
  }

  coords  <- st_coordinates(points_sf)
  x_seq   <- seq(-180, 180, by = res(template)[1])
  y_seq   <- seq(-90,   90, by = res(template)[2])
  grid    <- expand.grid(x = x_seq, y = y_seq)

  kde_values <- sapply(1:nrow(grid), function(i) {
    dx <- coords[, 1] - grid$x[i]
    dy <- coords[, 2] - grid$y[i]
    sum(exp(-(dx^2 + dy^2) / (2 * bandwidth^2)))
  })

  r <- rast(
    data.frame(x = grid$x, y = grid$y, value = kde_values),
    type = "xyz",
    crs = "EPSG:4326"
  )

  r <- r / nrow(points_sf)
  return(r)
}


kde_pre  <- compute_kde(filter(mentions_sf, period == "pre_crisis"),  raster_template, KDE_BANDWIDTH)
kde_peak <- compute_kde(filter(mentions_sf, period == "crisis_peak"), raster_template, KDE_BANDWIDTH)
kde_post <- compute_kde(filter(mentions_sf, period == "post_crisis"), raster_template, KDE_BANDWIDTH)

kde_pre_r  <- resample(kde_pre,  raster_template, method = "bilinear")
kde_peak_r <- resample(kde_peak, raster_template, method = "bilinear")
kde_post_r <- resample(kde_post, raster_template, method = "bilinear")

diff_crisis   <- kde_peak_r - kde_pre_r
diff_recovery <- kde_post_r - kde_peak_r

# Save rasters
writeRaster(kde_pre_r,     "../out/kde/kde_pre.tif",      overwrite = TRUE)
writeRaster(kde_peak_r,    "../out/kde/kde_peak.tif",     overwrite = TRUE)
writeRaster(kde_post_r,    "../out/kde/kde_post.tif",     overwrite = TRUE)
writeRaster(diff_crisis,   "../out/kde/diff_crisis.tif",  overwrite = TRUE)
writeRaster(diff_recovery, "../out/kde/diff_recovery.tif",overwrite = TRUE)
```

```{r}
#Read Rasters

library(terra)

kde_pre_r     <- rast("../out/kde/kde_pre.tif")
kde_peak_r    <- rast("../out/kde/kde_peak.tif")
kde_post_r    <- rast("../out/kde/kde_post.tif")
diff_crisis   <- rast("../out/kde/diff_crisis.tif")
diff_recovery <- rast("../out/kde/diff_recovery.tif")
```


#Define World polygons

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
```


# Define functions fot plotting

```{r}
raster_to_df <- function(r) {
  as.data.frame(r, xy = TRUE) %>%
    rename(value = 3) %>%
    filter(!is.na(value))
}

# Period labels for plot titles
period_labels <- c(
  pre_crisis   = "Pre-crisis",
  crisis_peak  = "Crisis peak",
  post_crisis  = "Post-crisis"
)

# Consistent centroid colours per period so they read the same across all plots
period_colours <- c(
  pre_crisis  = "#1a9641",
  crisis_peak = "#d7191c",
  post_crisis = "#2c7bb6"
)

# !! Centroid point size
CENTROID_SIZE <- 4

# Base KDE plot (single period, no centroid yet)

plot_kde_base <- function(raster_df,
                          title,
                          diverging = FALSE,
                          shared_limit = NULL,
                          low  = "#2166AC",   # darker blue
                          mid  = "#F7F7F7",   # near-white
                          high = "#B2182B") { # darker red

  if (diverging) {

    # Use shared limit if supplied so range becomes the same
    limit <- ifelse(
      is.null(shared_limit),
      max(abs(raster_df$value), na.rm = TRUE),
      shared_limit
    )

    # Clipping to avoid one extreme pixel dominating colours
    raster_df <- raster_df %>%
      mutate(value = pmax(pmin(value, limit), -limit))

    fill_scale <- scale_fill_gradient2(
  low       = low,
  mid       = mid,
  high      = high,
  midpoint  = 0,
  limits    = c(-limit, limit),
  oob       = scales::squish,

  # More granular legend ticks
  breaks = pretty(
    c(-limit, limit),
    n = 7
  ),

  # Consistent decimal formatting
  labels = scales::label_number(
    accuracy = 0.0001
  ),

  name = expression(Delta~density)
)

  } else {

    fill_scale <- scale_fill_gradient(
      low  = "white",
      high = "#B2182B",
      name = "Attention\ndensity"
    )
  }

  ggplot() +
    geom_raster(data = raster_df,
                aes(x = x, y = y, fill = value)) +

    geom_sf(data = world,
            fill = NA,
            color = "grey35",
            linewidth = 0.25) +

    fill_scale +

    coord_sf(
      xlim = MAP_XLIM,
      ylim = MAP_YLIM,
      expand = FALSE
    ) +
    annotation_scale(
    location      = "br",
    width_hint    = 0.2,
    style         = "bar",
    unit_category = "metric"
  ) + #include scale on bottom right

    labs(title = title) +

    theme_minimal(base_size = 11) +

    theme(
      axis.title      = element_blank(),
      axis.text       = element_blank(),
      axis.ticks      = element_blank(),
      panel.grid      = element_blank(),

      plot.title = element_text(
        size = 20
      ),

      legend.position = "right",

      legend.title = element_text(
        face = "bold"
      )
    )
}
```


# Add a single centroid point to an existing plot
```{r}
add_centroid <- function(p, centroid_row, colour, size = CENTROID_SIZE) {
  p +
    geom_point(
      data  = centroid_row,
      aes(x = lon, y = lat),
      color = colour,
      fill  = "white",
      shape = 21,       # filled circle with border
      size  = size,
      stroke = 1.5
    ) +
    geom_text(
      data  = centroid_row,
      aes(x = lon, y = lat, label = period_labels[as.character(period)]),
      vjust = -1, size = 3, color = colour, fontface = "bold"
    )
}
```


# Add all three centroids.
### Legacy code for connecting arrows. Is currently omitted, but naming is still there from earlier
```{r}
add_centroid_trajectory <- function(p, centroids_df, colours, size = CENTROID_SIZE) {

  segments <- centroids_df %>%
    mutate(
      lon_end = lead(lon),
      lat_end = lead(lat)
    ) %>%
    filter(!is.na(lon_end))

  p +
    geom_point(
      data  = centroids_df,
      aes(x = lon, y = lat, color = period),
      shape = 19,
      size  = size
    ) +
    scale_color_manual(
      values = colours,
      labels = period_labels,
      name   = "Period"
    ) +
    geom_text(
      data  = centroids_df,
      aes(x = lon, y = lat, label = period_labels[as.character(period)]),
      vjust = -1, size = 3, fontface = "bold",
      color = colours[as.character(centroids_df$period)]
    )
}
```



# Per-period KDE plots with single centroids
```{r}
centroid_pre  <- centroids %>% filter(period == "pre_crisis")  %>% mutate(period = period)
centroid_peak <- centroids %>% filter(period == "crisis_peak") %>% mutate(period = period)
centroid_post <- centroids %>% filter(period == "post_crisis") %>% mutate(period = period)

p_pre <- plot_kde_base(
  raster_to_df(kde_pre_r),
  period_labels["pre_crisis"]
) %>% add_centroid(centroid_pre,  period_colours["pre_crisis"])

p_peak <- plot_kde_base(
  raster_to_df(kde_peak_r),
  period_labels["crisis_peak"]
) %>% add_centroid(centroid_peak, period_colours["crisis_peak"])

p_post <- plot_kde_base(
  raster_to_df(kde_post_r),
  period_labels["post_crisis"]
) %>% add_centroid(centroid_post, period_colours["post_crisis"])
```


# Difference raster plots between periods
```{r}
# Shared scale for both difference maps

shared_limit <- quantile(
  c(
    abs(values(diff_crisis)),
    abs(values(diff_recovery))
  ),
  probs = 0.999,
  na.rm = TRUE
)

centroid_diff_crisis  <- centroids %>% filter(period != "post_crisis")  %>% mutate(period = period)

p_diff_crisis <- plot_kde_base(
  raster_to_df(diff_crisis),
  "Change in Attention: Crisis Peak vs Pre-Crisis",
  diverging = TRUE,
  shared_limit = shared_limit
) %>%
  add_centroid_trajectory(
    centroid_diff_crisis,
    period_colours
  )


centroid_diff_recovery  <- centroids %>% filter(period != "pre_crisis")  %>% mutate(period = period)

p_diff_recovery <- plot_kde_base(
  raster_to_df(diff_recovery),
  "Change in Attention: Post-Crisis vs Crisis Peak",
  diverging = TRUE,
  shared_limit = shared_limit
) %>%
  add_centroid_trajectory(
    centroid_diff_recovery,
    period_colours
  )
```


# Combine relevant figures

```{r}
# Three-period overview with individual centroids
p_overview <- p_pre / p_peak / p_post +
  plot_annotation(
    title    = "Geographic attention in Danish news coverage by period",
    subtitle = "KDE density surface with weighted mention centroid (circle)",
    theme    = theme(plot.title = element_text(size = 13, face = "bold"))
  )

# Difference maps with full trajectory
p_diffs <- p_diff_crisis | p_diff_recovery +
  plot_annotation(
    title    = "Shifts in geographic attention: raster difference + centroid trajectory",
    subtitle = "Red = gained attention, blue = lost. Arrows show centroid movement pre → peak → post",
    theme    = theme(plot.title = element_text(size = 13, face = "bold"))
  )
```


# Save

```{r}
ggsave("../out/kde/kde_centroid_overview.png", p_overview,
       width = 10, height = 16, dpi = 300)

ggsave("../out/kde/kde_centroid_diffs.png", p_diffs,
       width = 16, height = 7, dpi = 300)

ggsave("../out/kde/kde_centroid_pre.png",  p_pre,  width = 10, height = 6, dpi = 300)
ggsave("../out/kde/kde_centroid_peak.png", p_peak, width = 10, height = 6, dpi = 300)
ggsave("../out/kde/kde_centroid_post.png", p_post, width = 10, height = 6, dpi = 300)

ggsave("../out/kde/kde_diff_crisis_centroid.png",   p_diff_crisis,   width = 10, height = 6, dpi = 300)
ggsave("../out/kde/kde_diff_recovery_centroid.png",  p_diff_recovery, width = 10, height = 6, dpi = 300)

```


```{r}
cat("\nCentroid coordinates:\n")
print(centroids)
```

